\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-29 chapters/conclusion.tex
Writing patient paper

>>>>

very small group of interest, but available right now, no implantation needed,

even if it is the most suited strategy for a patient, it will probably be
improved if a combination of eye tracking, residual motor control and other
biosensors are used.

true power in the future is probably focused on
invasive, active paradigms if recording technology and surgical methods
improve lowering the risk. orders of magnitude higher itr and more natural
communication.
characterized by publication date, but might indicate it is of interest to
heavily focus this research now on applied 'vraagstukken' with patients

### 2024-09-29 chapters/patients.tex
Writing patient paper

Communication BCIs have a target population consisting of patients in various
stages of paralysis or \ac{lis}, often suffering from oculomotor impairments.
This reduces their performance operating visual oddball \ac{bci}s, since they
are presumed unable to comfortably redirect their gaze at the desired target,
i.e., perform overt \ac{vsa}.
They might more naturally operate in covert \ac{vsa}, where the gaze and
\ac{vsa} do not coincide.
Several studies show that performance drops in covert VSA \todo{[2, 1], our
study}, necessitating gaze-independent solutions.
Our previous study, presented in chapter \todo{ref chapter}, indicates gaze-independent performance can be
improved in healthy subjects using a suited decoding strategy.

These studies build on the assumption that patients with eye motor impairment
would feel comfortable operating
Visual paradigms

Need for verification in patients, cite different characteristics, differences
in setting, environment and equipment,
\todo{goal is class 1, but challenging to perform research experiments and to
recruit, cite wolpaw}

problems eye tracking

Literature for visual \ac{bci} in eye motor impaired patients is sparse.
\todo{List conclusions of Peters2020 (severely eye motor impaired als patients
able to use a hex-o-spell)}

Therefore, we aim to apply the concepts from earlier work an literature to
patients with motor impairment and, in addition, eye motor impairment.
The objectives of this case study are as follows:
\begin{enumerate*}
  \item Explore capabilities and experienced comfort of eye motor impaired
    patients,
  \item evaluate the performance of a gaze-independent visual \ac{bci} in patients
  \item verify if this performance can be improved with a suitable decoding
    strategy.
\end{enumerate*}

\section{Materials \& methods}
\subsection{Patient recruitment}
Patients were recruited from across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France) and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Patients were recruited based on the following criteria.
To qualify for inclusion, patients must
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according the \ac{bci}
	patient selection criteria presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \todo{specific cutoff based on vision}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Patients were excluded if they
\begin{enumerate}
  \item have a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw}
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item have a predisposition to or have a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item have a severe loss in vision or hearing, that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the out-
        come.\label{item:patients/exclusion/cognitive}
  \item be able to understand the experiment instructions and cooperate,
  \item have any other limitations preventing them from performing the given task.
\end{enumerate}

@@

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table.tex}
  \caption{Presentation of included patients including their diagnosis and
  capabilities. (Trach.: patient underwent a tracheotomy, Cls.: classification according to~\textcite{Wolpaw2006}).
  }
  \label{tab:patients/patients}
\end{table}

@@

To adapt the interface for the patient setting, the number of blocks was
decreased to 6 per \ac{vsa} setting.
An additional \ac{free \ac{vsa}} condition was introduced.
Here, the patient was instructed to perform the task as they deemed most
comfortable.
The cued \emph{split attention} setting proposed
by~\textcite{VanDenKerchove2024} is not studied here, since we are interested
in natural \ac{vsa} operation settings for gaze-impaired patients.
If the patient was not fully paralyzed, they were instructed not to move the head.
\Ac{isi} was increased to\tod{number} to decrease task difficulty.
The experiment alsy started with a training block in each condition, where the
patient was instructed with feedback on their performance to make sure they
understood and were able to perform the task.

>>>>

Communication BCIs have a target population consisting of patients in various
stages of paralysis or \ac{lis}, often suffering from oculomotor impairments.
This reduces their performance operating visual oddball \ac{bci}s (see
section\todo{ref section} for an overview), since they
are presumed unable to comfortably redirect their gaze at the desired target,
i.e., perform overt \ac{vsa}.\todo{cite some visual gaze-independent bci}
They might more naturally operate in covert \ac{vsa}, where the gaze and
\ac{vsa} do not coincide.
Several studies show that performance drops when not fixating the intended
target~\cite{Brunner2010, Frenzel2011,
Treder2010, RonAngevin2019, VanDenKerchove2024} \todo{[2, 1], our
study}, necessitating gaze-independent solutions.

Usually, they aim to improve gaze-independent \ac{bci} by optimizing
performance in cued covert \ac{vsa}.
These studies build on the assumption that patients with eye motor impairment
would feel comfortable operating an interface in pure covert \ac{vsa} with
central fixation.
One could argue that a \ac{bci} that is only verified to work when a central
fixation is maintained, could also be considered gaze-dependent.
This does not account for the residual eye motor capabilities of most severly
paralyzed or locked-in patients and the comfort they experience while
performing the gaze fixation.

Our previous study, presented in chapter \todo{ref chapter}  partially tried to account for this
and showed gaze-independent performance can be improved
improved in healthy subjects using a suited decoding strategy.
Yet, there is a strong need for verification of results obtained in healthy
samples.
It is a striking constatation that studies reporting on
gaze-independent visual \ac{bci} in patients that are actually eye-motor
impaired are very few.
in patients, cite different characteristics, differences
in setting, environment and equipment,
\todo{List conclusions of Peters2020 (severely eye motor impaired als patients
able to use a hex-o-spell)}

Eventually, one of the end goals of this research line is to develop
gaze-independent \ac{bci} for patients that
are fully locked-in and have no option left than to use a \ac{bci}.
However, this group is very small and it is often a challenge to recruit them
into a study and perform experiments with them~\cite{Wolpaw2006}.
Patients with less severe paralysis or in less progressed disease stages that struggle with
eye-tracking technology could also benefit from
solutions tailored to their specific situation.

Therefore, we aim to apply the concepts from earlier work an literature to
patients with various degrees motor impairment and various
degrees of eye motor impairment.
The objectives of this case study are as follows:
\begin{enumerate*}
  \item Explore capabilities and experienced comfort of eye motor impaired
    patients,
  \item evaluate the performance of a gaze-independent visual \ac{bci} in patients
  \item verify if this performance can be improved with a suitable decoding
    strategy.
\end{enumerate*}

\section{Materials \& methods}
\subsection{Patient recruitment}
Patients were recruited from across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France) and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Patients were recruited based on the following criteria.
To qualify for inclusion, patients must
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according the \ac{bci}	patient selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Patients were excluded if they
\begin{enumerate}
  \item have a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw}
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item have a predisposition to or have a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item have a severe loss in vision or hearing, that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the out-
        come.\label{item:patients/exclusion/cognitive}
  \item be able to understand the experiment instructions and cooperate,
  \item have any other limitations preventing them from performing the given task.
\end{enumerate}

@@

\todo{check Usability and Workload of Access Technology for People With Severe
Motor Impairment: A Comparison of Brain-Computer Interfacing and Eye Tracking}

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table.tex}
  \caption{Presentation of included patients including their diagnosis and
  capabilities. (Trach.: patient underwent a tracheotomy, Cls.: classification according to~\textcite{Wolpaw2006}).
  }
  \label{tab:patients/patients}
\end{table}
\todo{Replace with classification according to 40. Kübler A, Birbaumer N.
Brain-computer interfaces and communication in paralysis: extinction of goal
directed thinking in completely paralysed patients? Clin Neurophysiol.
2008;119:2658-2666.}
\todo{include year of diagnosis}

@@

To adapt the interface for the patient setting, the number of blocks was
decreased to 6 per \ac{vsa} setting.
An additional \emph{free \ac{vsa}} condition was introduced.
Here, the patient was instructed to perform the task as they deemed most
comfortable.
The cued \emph{split attention} setting proposed
by~\textcite{VanDenKerchove2024} is not studied here, since we are interested
in natural \ac{vsa} operation settings for gaze-impaired patients.
If the patient was not fully paralyzed, they were instructed not to move the head.
\Ac{isi} was increased to\todo{number} to decrease task difficulty.
The experiment alsy started with a training block in each condition, where the
patient was instructed with feedback on their performance to make sure they
understood and were able to perform the task.
